---
title: "The URPS demand model, end to end"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The URPS demand model, end to end}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Chunks are not evaluated at build time: they call the full demand stack and are
# for illustration. The coefficient tables are placeholders
# (calibration_status = "placeholder_uncalibrated"), so any numbers here are
# structure, not results. See docs/DEMAND_METHODS.md.
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
library(urpssim)
library(dplyr)
```

This vignette walks the demand side of `urpssim` from the obstetric life course to
required provider FTE, the dynamic disease model, calibration, and the versioned
demand contract that feeds downstream repositories. It mirrors the methods in
`docs/DEMAND_METHODS.md`.

## 1. A synthetic female population

```{r pop}
pop_by_age_year <- tidyr::expand_grid(year = 2025:2050, age = 40:85) |>
  dplyr::mutate(population = round(2e6 * exp(-0.02 * (age - 40))))
```

## 2. Life-course demand → service volumes → FTE

The pathway runs childbirth exposure → pelvic-floor conditions → care-seeking →
service volumes, then hands the volumes to the work-RVU FTE conversion (`R/17`).
`wrvu_per_fte` is illustrative here; in practice derive it with
`calibrate_wrvu_per_fte()`.

```{r fte}
fte <- lifecourse_required_fte(
  pop_by_age_year,
  wrvu_per_fte = WRVU_PER_FTE_BENCHMARK[["median"]],
  scenario = "baseline", n = 5e4, seed = 1
)
fte$required_fte
```

Scenarios shift the primary exposure or the care pathway (BMI reduction lives in
the prevention scenario, not at the model's center):

```{r scenarios}
lifecourse_required_fte(pop_by_age_year, WRVU_PER_FTE_BENCHMARK[["median"]],
                        scenario = "delivery_mode", cesarean_rate = 0.60, seed = 1)
lifecourse_required_fte(pop_by_age_year, WRVU_PER_FTE_BENCHMARK[["median"]],
                        scenario = "reduced_barriers", seed = 1)
```

## 3. Parameter-uncertainty intervals

```{r uncertainty}
ci <- lifecourse_demand_trajectory_ci(pop_by_age_year, n = 2e4, n_draws = 200, seed = 1)
ci  # median + lo/hi for care-seeking and service units
```

## 4. Dynamic multistate disease model (DMDM)

Longitudinal onset/remission/death; the open engine can be reweighted to Census
projections so counts match official demography.

```{r dmdm}
tr <- dmdm_open_prevalence_trajectory(pop_by_age_year, 2025, 2050,
                                      reweight_to_projection = TRUE, seed = 1)
tr  # year, population, prev_ui/pop/ai, incidence
```

Fit the hazards from a longitudinal panel (SWAN) instead of the placeholders:

```{r fit}
td  <- dmdm_transition_data(swan_panel)          # your SWAN-style person-year panel
fit <- fit_dmdm_transitions(td)                   # status = "fitted"
dmdm_open_prevalence_trajectory(pop_by_age_year, 2025, 2050, transitions = fit)
```

Prolapse is graded (POP-Q 0–4) and regresses, unlike incontinence. Until a POP
panel is fitted, use the cited literature transitions (`derived_by_analogy`);
with a staged panel, fit the per-stage progression/regression hazards directly:

```{r pop_staged}
tr_pop <- dmdm_transitions_with_pop_literature()  # MOAD/WHI/SWEPOP-derived POP row
tr_pop$pop_progression                             # annual stage k -> k+1
tr_pop$pop_regression                              # annual stage k -> k-1 (mild POP resolves)

td_st <- dmdm_transition_data(swan_panel, stage_cols = c(pop = "pop_stage"))
fit_dmdm_transitions(td_st, stage_conditions = "pop")  # + pop_progression/pop_regression
```

## 5. Calibration and back-test

```{r calibration}
obs <- tibble::tibble(category = c("urps_office_visits", "sling_procedure_volume",
                                   "prolapse_procedure_volume"),
                      observed = c(3.6e6, 1.0e5, 1.0e5))   # HCUP SASD / Medicare / NAMCS
cal <- calibrate_lifecourse_demand(fte$service_volumes, obs)
cal$scalars   # scalar = observed / predicted (HDMM Exhibit 11)
```

## 6. Concordance across demand estimands

```{r concordance}
bands <- example_female_population_by_band()
demand_long <- dplyr::bind_rows(
  compute_demand_denominators(bands),                 # D1/D2/D3
  compute_demand_denominators_lifecourse(bands),      # D4 (obstetric-weighted)
  lifecourse_demand_estimand(                          # D5 (life-course service)
    lifecourse_demand_trajectory(pop_by_age_year, n = 2e4, seed = 1)$demand_summary))
supply <- tibble::tibble(year = sort(unique(bands$year)),
                         effective_fte_median = seq(1300, 1500,
                           length.out = length(unique(bands$year))))
assess_demand_concordance(compute_growth_adequacy(supply, demand_long), demand_long)
```

## 7. Geographic (isochrone) demand

```{r geo}
geo_need <- data.frame(geo_id = 1:4, need = c(100, 400, 150, 100),
                       nearest_provider_min = c(15, 90, 200, 240),
                       access_ratio = c(3, 0.5, 0.2, 0.1))
geographic_demand_summary(geo_need)  # need within 30/60/120/180 min; need-weighted access
```

## 8. Versioned demand contract (feeds cliff / twostep / isochrones)

Both exporters refuse to write a contract built on uncalibrated transitions. Every
transition set shipped with the package is still a placeholder, so these calls
pass `allow_uncalibrated = TRUE` to declare the artifact exploratory. Drop that
argument once the hazards are fitted (see `fit_dmdm_transitions()`) and the
export becomes publication-facing on its own.

```{r contract}
traj <- lifecourse_demand_trajectory(pop_by_age_year, n = 2e4, seed = 1)
# tiers 5-6
export_hdmm_demand_contract(traj$demand_summary, output_directory = tempdir(),
                            allow_uncalibrated = TRUE)
# tier 3 + per-condition
export_dmdm_demand_contract(tr, output_directory = tempdir(),
                            allow_uncalibrated = TRUE)
```
