---
title: "Getting started with urpssim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with urpssim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(urpssim)
```

## What this package answers

Will the US supply of urogynecology and reconstructive pelvic surgery (URPS)
providers keep pace with demand, and where will access fall short?

It answers that by projecting supply and demand **in the same units** —
full-time-equivalent providers — and reporting the gap with an uncertainty
interval. Supply comes from an individual-level microsimulation of a provider
cohort; demand comes from age-banded population series converted to required FTE
through a work-RVU workload basket.

The rest of this vignette runs the model once and then explains the three things
newcomers reliably trip over.

## A first run

`run_workforce_microsimulation()` is the entry point. It takes many arguments;
almost all have defaults. The two you cannot skip are a **base-year gap** and,
if any input is analogy-derived, an explicit acknowledgement that you know it is.

```{r first-run}
gap <- baseline_gap(
  base_supply_fte    = 120,
  adequacy           = 0.95,
  method             = "assumed",
  calibration_status = "uncalibrated_illustrative",
  evidence           = "illustrative example, not a published estimate"
)

result <- run_workforce_microsimulation(
  baseline_supply           = 120,
  use_certification_cohorts = FALSE,
  years                     = 2025:2030,
  n_iterations              = 5,
  supply_scenarios          = supply_scenario_registry(20)["baseline"],
  pop_by_band               = example_female_population_by_band(2025:2030),
  baseline_gap_estimate     = gap,
  allow_analogy             = TRUE,
  verbose                   = FALSE
)

names(result)
```

`n_iterations = 5` keeps this vignette fast. Real runs use 200.

### Reading the result

`supply` is the projected provider FTE by year and scenario:

```{r supply}
head(result$supply[, c("year", "scenario_label", "effective_fte_median")])
```

`fte_gap` is the headline: required FTE against projected supply, in the same
units, so subtracting them is meaningful.

```{r gap}
head(result$fte_gap)
```

### What the run says about itself

Every run carries metadata describing how much you should trust it. This is not
decoration — it is the part most likely to change your interpretation.

```{r meta}
meta <- result$scenario_meta
meta$demand_calibrated
meta$example_only
```

`demand_calibrated` is `FALSE` here because no independent anchor was supplied.
The projection still ran; it simply is not anchored to an observed national
quantity, and the object says so rather than letting you assume otherwise.

## Three things that will surprise you

### 1. An interval is not automatically a forecast interval

If you do not pass a `supply_parameter_spec()`, every Monte Carlo iteration uses
the *same* entrant rate. The resulting band describes individual stochasticity
only — the luck of who retires when — and not uncertainty about the future.

In the 2020→2023 back-test, intervals of that kind were 0–40 providers wide on a
count near 1,300 (two arms had **zero** width) and covered the observed value in
0 of 8 arms. Drawing the entrant rate per iteration widens them to 129–148.

```{r spec}
spec <- supply_parameter_spec(
  entrant_series = c(50, 57, 53, 59, 59),
  entrant_mean   = 55
)
spec$quantified
```

Pass that as `parameter_spec` and the interval starts carrying forecast
uncertainty. Without it, report the band as a Monte Carlo range —
`interval_label()` will not let you call it a prediction interval.

### 2. Every input declares a calibration tier

`baseline_gap()` requires `calibration_status`; it will not infer one. The same
arithmetic is *calibrated* when it comes from a survey fielded on
urogynecologists and *derived by analogy* when it comes from another specialty's
published distribution, and the function refuses to guess which it was handed.

```{r tiers, error = TRUE}
# Omitting the tier for a non-'assumed' method is an error, not a default.
baseline_gap(1306, 0.95, method = "capacity_survey", evidence = "x")
```

This matters because the base-year shortfall passes through to the headline gap
**with a coefficient of one**. A borrowed number must not be reportable as a
measurement.

### 3. Some things are deliberately not resolved

The package reports its own open items rather than burying them.

```{r status}
capacity_status()$resolved
```

The base-year capacity anchor is a published physical-therapy distribution
standing in for a URPS practice-capacity survey that has not been fielded. It is
the single highest-value missing input.

A useful structural fact: because
`required_fte(t) = anchor × wRVU(t) / wRVU(base)`, the delegation matrix, the
demand calibration and the workload *levels* all cancel. A 2.1× change in
calibration moved 2050 required FTE by 0.25%. Only the anchor and the *shape* of
demand growth move the answer — which is why the anchor is the thing to fix.

## Finding your way around

There are over 400 exported objects. Two roxygen tags make them navigable:

* **`@concept`** names the layer, so `help.search("geography", package = "urpssim")`
  narrows to it. The eight layers match the `R/` filename prefixes: `supply`,
  `demand`, `geography`, `calibration`, `validation`, `reporting`, `core`, `data`.
* **`@family`** names the module, so every help page lists its siblings under
  "See also". Start anywhere and walk sideways.

For the code itself, `docs/ARCHITECTURE.md` maps the package and
`docs/GUARDS.md` explains each guard and the defect that motivated it. Both live
in the repository rather than the installed package.

## Where to go next

* `vignette("demand-model")` — how population becomes required FTE
* `?run_workforce_microsimulation` — every argument
* `?baseline_gap` — stating the base-year shortfall honestly
* `backtest_status()` — how the engine scored against a held-out period
