---
title: "urpssim: supply, demand, access, gap, and sensitivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{urpssim: supply, demand, access, gap, and sensitivity}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
library(urpssim)
suppressPackageStartupMessages({library(dplyr); library(tibble)})
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
has_sf     <- requireNamespace("sf", quietly = TRUE)
```

This vignette walks the whole model — **supply → demand → access → gap →
sensitivity** — end to end. It deliberately uses the pieces that run without the
`mufflyaccess` contract, so it reproduces anywhere. The full, contract-backed
entry point is `run_workforce_microsimulation()`.

## 1. Supply

Each provider is an agent with an age, an entry cohort, and an age-band
retirement hazard; supply is the Monte-Carlo distribution of active effective
FTE. Empirical URPS hazards can be swapped in via `retirement_source`.

```{r supply}
set.seed(20260801)
agents <- initialize_provider_agents(1306, "FPMRS", 2025)
agents$sex <- rep(c("female", "male"), length.out = nrow(agents))

supply <- run_supply_microsimulation(
  agents, years = 2025:2040, entrants_per_year = 70,
  n_iterations = 40, verbose = FALSE
)$summary
tail(supply[c("year", "headcount_median", "effective_fte_median")], 3)
```

## 2. Demand

Demand carries three independent estimands (prevalent PFD cases, consultations,
surgical volume), each with its own age profile, so they diverge as the
population ages. Here we use the illustrative age-banded population; the real
Census-NPP series loads through `resolve_demand_population()`.

```{r demand}
pop <- example_female_population_by_band(2025:2040)
demand <- compute_demand_denominators(pop)
demand %>% filter(year == 2040) %>% select(estimand, label, demand_cases)
```

## 3. Access

The E2SFCA engine runs on the real 83k-tract female-65+ demand. Provider
catchments here are illustrative buffers around two cities (the real national run
consumes Valhalla drive-time polygons via `load_provider_isochrones()`).

```{r access, eval = has_sf}
provs <- sf::st_as_sf(
  data.frame(coord_id = c("denver", "nyc"), drive_time = c(60L, 60L),
             lon = c(-104.99, -74.00), lat = c(39.74, 40.71)),
  coords = c("lon", "lat"), crs = 4326)
iso <- suppressWarnings(sf::st_buffer(provs, 120000))
acc <- workforce_access_summary(iso, tibble(provider_id = c("denver", "nyc"),
                                            supply = c(8, 20)))
sprintf("%.1f%% of women fall in an access desert (%d tracts)",
        acc$access_desert_share_pct, acc$n_tracts)
```

```{r access-fig, eval = has_sf && has_ggplot}
fig_access_threshold_shares(acc$threshold_shares)
```

## 4. Gap and outlook

Replacement ratio = entrants / expected departures, classified on the cliff
cutpoints (Adequate ≥ 1.2, Marginal 0.8–1.2, Insufficient < 0.8).

```{r outlook}
outlook <- tibble(
  scenario_label = c("Status quo", "Enhanced training", "70% conversion"),
  entrants = c(70, 77, 49), departures = 57) %>%
  mutate(replacement_ratio = entrants / departures,
         outlook = classify_workforce_outlook(replacement_ratio))
outlook %>% select(scenario_label, replacement_ratio, outlook)
```

```{r outlook-fig, eval = has_ggplot}
fig_replacement_outlook(outlook)
```

## 5. Sensitivity

A global probabilistic sensitivity analysis samples every uncertain input
jointly and ranks which ones drive the spread of the 2050 gap (PRCC tornado).
The workforce evaluator is `psa_workforce_gap()`; here we show the engine on a
compact synthetic gap model.

```{r psa}
inputs <- list(
  psa_uniform("entrants", 45, 90),
  psa_discrete("retirement_source", c("hwsm", "urps_empirical")),
  psa_uniform("base_adequacy", 0.85, 1.02),
  psa_discrete("population_series", c("mid", "low", "hi")))
evaluate <- function(p) {
  dem <- switch(p$population_series, low = -2, mid = 0, hi = 2)
  ret <- if (p$retirement_source == "urps_empirical") -1.5 else 0
  -(1 - p$base_adequacy) * 100 - 0.15 * (p$entrants - 67) + dem - ret
}
psa <- run_psa(inputs, evaluate, n = 300, seed = 1, verbose = FALSE)
psa_tornado(psa)
```

```{r psa-fig, eval = has_ggplot}
plot_psa_tornado(psa_tornado(psa))
```

## Putting it together

`workforce_summary_report()` folds a full `run_workforce_microsimulation()`
result — the final-year FTE gap, the demand-concordance verdict, the replacement
outlook, the calibration tier of every input, and (when isochrones are supplied)
the access-desert share — into one PI-facing object with a `print()` method.
```
